Co-moving stars in \textsl{Gaia DR1}:\\ An abundance of very wide separation co-moving pairs

### 2017-04-04 paper/ms.tex
clean up appendix

If the velocity prior $p(\vec{v})$ is also Gaussian, the integrals over velocity
in both cases can be performed analytically:
We use a mixture of three isotropic, zero-mean Gaussian distributions
\begin{equation}
  p(\vec{v}) = \sum_{m=1}^3 \, w_m \, \mathcal{N}(0, \sigma_{v,m}^2)
\end{equation}
with velocity dispersions $(\sigma_{v,1}, \sigma_{v,2}, \sigma_{v,3}) = (15, 30, 50)
~\kms$ and weights $(w_1,w_2,w_3) = (0.3, 0.55, 0.15)$
meant to encompass young thin disk stars to halo stars.
These numbers are empirically chosen to account for the distribution of
velocities of the \tgas\ stars.
We derive the relevant expressions in Appendix~\ref{sec:appendix}.
After marginalizing over velocity, the likelihood integrands only depend on
distance; we numerically compute the integrals over the true distances of each
star in a pair using Monte Carlo integration with $K$ samples from
the distance posterior.
\begin{equation}
  \int \, \dd r \, \tilde{L}(r) \, p(r \given \varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(r_k)
\end{equation}
where $\tilde{L}(r)$ is the velocity-marginalized likelihood function.
In order to generate a sample of distances from the distance posterior $p(r_i \given \varpi_i)$,
we need to assume a distance prior. We adopt the uniform density prior
\citep{Bailer-Jones:2015aa} with a maximum distance of 1~kpc.
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.

@@

We have also found a large number of very wide separation ($>1$~pc)
mutually exclusive co-moving pairs, in which the stars are the only co-moving neighbor
of each other and not part of large connected components.
These are most likely remnants of dissolving wide binaries (\citealt{Jiang:2010aa}).
The abundance of highly probable wide separation co-moving pairs
conclusively shows that there is no strict cut-off semi-major axis for wide binary systems
(\citealt{Wasserman:1987aa}).
The presence of these pairs and similar separation distribution have already
been noticed by \citealt{Shaya:2011aa} using the \project{Hipparcos} data.
If confirmed with radial velocity measurements, this population should still be
relatively young compared to the general disk field population.
Modeling the color-magnitude diagram distribution of these stars can shed some
light on this issue.
If they are remnants of dissolving systems that were born coeval,
the sample of very wide separation co-moving pairs can potentially be used
to measure the recent ($\lesssim 1$~Gyr)
star formation history in the Solar neighborhood.
Co-moving stars with separation less than 1~pc are very promising candidates
for wide binaries. They are found to be pairs of stars of varying stellar types.
Some of these pairs, such as main sequence-red giant or F/G/K-M dwarfs,
will be particularly valuable for testing theoretical
stellar models and calibrating observational measurements of low mass stars.

@@

In what follows, all vectors are column vectors, unless we have transposed them.
A relevant exponential integral solution is
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,
    \transp{[\vec{x}-\vec{\nu}]} \,
    \inv{\mat{A}} \,
    [\vec{x}-\vec{\nu}] - \Delta) \, \dd \vec{x}\right]
  &=& +\frac{1}{2}\ln ||2\pi\,\mat{A}|| -\Delta
  \quad , \label{eq:gauss-int}
\end{eqnarray}
where $\vec{x}$ and $\vec{\nu}$ are $D$-dimensional vectors, $\mat{A}$ is a
positive definite matrix, $\Delta$ is a scalar, and the integral is over all of
$D$-dimensional $\vec{x}$-space.
To cast our problem in this form, we will need to complete the square of the
exponential argument.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,\transp{[\vec{x}-\vec{\nu}]}\,\inv{\mat{A}}\,[\vec{x}-\vec{\nu}] + \Delta
  &=& \frac{1}{2}\,\transp{\vec{x}}\,\inv{\mat{A}}\,\vec{x} + \transp{\vec{x}}\,\mat{B}\,\vec{b} + C
  \quad ,
\end{eqnarray}
where $\mat{B}\,\vec{b}$ is an $D$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \vec{\nu} &=& -\mat{A}\,\mat{B}\,\vec{b}
  \\
  \Delta & = & C - \frac{1}{2}\,\transp{\vec{\nu}}\,\inv{\mat{A}}\,\vec{\nu}
  \quad .
\end{eqnarray}
We will identify terms in our likelihood functions with $\mat{A}$,
$\mat{B}\,\vec{b}$, and $C$, convert to $\vec{\nu}$ and $\Delta$ and compute
the marginalized likelihood using \eqname~\ref{eq:gauss-int}.

@@

At given distance $r$, the velocity-marginalized likelihood can be computed
analytically using the expressions in Appendix~\ref{sec:appendixA}.
As described above, we use an isotropic, mixture-of-Gaussians prior on velocity
where the velocity dispersion of a given mixture component is $\sigma_{v,m}$ and
the 3-space variance tensor is $\mat{V}_m = \sigma_{v,m}^2 \, \eye$.
The likelihood for the data is also a Gaussian, as shown in
\eqname~\ref{eq:likefn}.
We will start by writing down expressions for the the likelihood multiplied by
the prior pdf for the velocities.
Here we slightly change the notation used in \sectionname~\ref{sec:methods} for
simplicity.

We construct a velocity-space data vector $y$ as follows:
\begin{equation}
  \vec{y} =
    \transp{\left(
      \begin{array}{c c c c}
        r_i\,\mu_{\alpha,i} &
        r_i\,\mu_{\delta,i} &
        r_j\,\mu_{\alpha,j} &
        r_j\,\mu_{\delta,j}
      \end{array}
    \right)}
\end{equation}
where again the subscripts $i,j$ refer to the indices of each star in the pair
and we have multiplied the observables (the proper motions) by the distances
$r_i, r_j$, which is permitted because we are conditioning on the distances.
Fundamentally, our hypothesis 1 model (the stars have the same velocity with a
small difference) is
\begin{equation}
  \vec{y} = \mat{M} \, \vec{v} + \mathrm{noise}
\end{equation}
where now the $4 \times 3$ transformation matrix $\mat{M}$ is a stack of the
transformation matrices for each star computed from the pair of sky positions
and using \eqname~\ref{eq:transformation}.
The noise (in the $y$ vector) is drawn from a $4 \times 4$ Gaussian with
block-diagonal covariance matrix, $\mat{\Sigma}$, constructed from the
covariance matrix of the observables, $\mat{C}_i, \mat{C}_j$, and the distances
$r_i, r_j$:
\begin{equation}
  \mat{\Sigma} = \left(
    \begin{array}{c c}
      r_i^2 \, \mat{C}_i & 0 \\
      0 & r_j^2 \, \mat{C}_j
    \end{array}
  \right)
\end{equation}

Given all these definitions, the likelihood function for hypothesis 1 is
\begin{eqnarray}
  p(\data \given \vec{v}, r_i, r_j) &=& r_i^2\,r_j^2\,
    \normal(\vec{y} \given \mat{M}\,\vec{v}, \mat{\Sigma}) \\
  \ln p(\data \given \vec{v}, r_i, r_j) &=& 2\,\ln r_i + 2\,\ln r_j
    -\frac{1}{2}\,\ln||2\pi\,\mat{\Sigma}|| \nonumber \\
    && \quad -\frac{1}{2}\,\transp{[\vec{y}-\mat{M}\,\vec{v}]}\,
      \inv{\mat{\Sigma}}\,
      [\vec{y}-\mat{M}\,\vec{v}]
  \quad ,
\end{eqnarray}
where the factor of $r_i^2\,r_j^2$ converts to units of one over \gaia\ data
from units of one over $y$, which includes the distances.

Now we multiply this likelihood with the velocity prior (here shown only for one
prior velocity mixture component) and complete the square.
In the original parameterization, we find:
\begin{eqnarray}
  \mat{A} &=& \inv{[\transp{\mat{M}}\,\inv{\mat{\Sigma}}\,\mat{M}+\inv{\mat{V}_m}]}
  \\
  \vec{\nu} &=& -\mat{A}\,\transp{\mat{M}}\,\inv{\mat{\Sigma}}\,\vec{y}
  \\
  \Delta &=& -2\,\ln r_i -2\,\ln r_j
    +\frac{1}{2}\,\ln||2\pi\,\mat{\Sigma}|| +\frac{1}{2}\,\ln||2\pi\,\mat{V}_m|| \nonumber \\
    && \quad +\frac{1}{2}\,\transp{\vec{y}}\,\inv{\mat{\Sigma}}\,\vec{y} -\frac{1}{2}\,\transp{\vec{\nu}}\,\inv{\mat{A}}\,\vec{\nu}
  \quad ,
\end{eqnarray}
which we plug in to \eqname~\ref{eq:gauss-int} to get the marginalized
likelihood conditioned on the two distances $r_i, r_j$.

The marginalized likelihood for the hypothesis 2 model (the stars have
independent velocities) is very similar.
In this case, the marginalized likelihood is a product of two independent
integrals $Q$, composed in the same way as the hypothesis 1 model but now for
each star individually, where
\begin{eqnarray}
  \vec{y} &=&
    \transp{\left(
      \begin{array}{c c}
        r\,\mu_{\alpha} &
        r\,\mu_{\delta}
      \end{array}
    \right)}
  \\
  \mat{\Sigma} &=& r^2 \, \mat{C}
  \\
\end{eqnarray}
and $\mat{M}$ is now the transformation matrix for one star. Then, again,
\begin{eqnarray}
  \mat{A} &=& \inv{[\transp{\mat{M}}\,\inv{\mat{\Sigma}}\,\mat{M}+\inv{\mat{V}_m}]}
  \\
  \vec{\nu} &=& -\mat{A}\,\transp{\mat{M}}\,\inv{\mat{\Sigma}}\,\vec{y}
  \\
  \Delta &=& -2\,\ln r_i -2\,\ln r_j
    +\frac{1}{2}\,\ln||2\pi\,\mat{\Sigma}|| +\frac{1}{2}\,\ln||2\pi\,\mat{V}_m|| \nonumber \\
    && \quad +\frac{1}{2}\,\transp{\vec{y}}\,\inv{\mat{\Sigma}}\,\vec{y} -\frac{1}{2}\,\transp{\vec{\nu}}\,\inv{\mat{A}}\,\vec{\nu}
  \quad ,
\end{eqnarray}
and
\begin{equation}
  Q = \frac{1}{2}\ln ||2\pi\,\mat{A}|| -\Delta \quad .
\end{equation}

>>>>

If the velocity prior $p(\vec{v})$ is also Gaussian, the integrals over velocity
in both cases can be performed analytically:
We use a mixture of three isotropic, zero-mean Gaussian distributions
\begin{equation}
  p(\vec{v}) = \sum_{m=1}^3 \, w_m \, \mathcal{N}(0, \sigma_{v,m}^2)
  \label{eq:vprior}
\end{equation}
with velocity dispersions $(\sigma_{v,1}, \sigma_{v,2}, \sigma_{v,3}) = (15, 30, 50)
~\kms$ and weights $(w_1,w_2,w_3) = (0.3, 0.55, 0.15)$
meant to encompass young thin disk stars to halo stars.
These numbers are empirically chosen to account for the distribution of
velocities of the \tgas\ stars.
We derive the relevant expressions in Appendix~\ref{sec:appendix}.
After marginalizing over velocity, the likelihood integrands only depend on
distance; we numerically compute the integrals over the true distances of each
star in a pair using Monte Carlo integration with $K$ samples from
the distance posterior.
\begin{equation}
  \int \, \dd r \, \tilde{L}(r) \, p(r \given \varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(r_k)
\end{equation}
where $\tilde{L}(r)$ is the velocity-marginalized likelihood function.
In order to generate a sample of distances from the distance posterior $p(r_i \given \varpi_i)$,
we need to assume a distance prior. We adopt the uniform density prior
\citep{Bailer-Jones:2015aa} with a maximum distance of 1~kpc.
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.

@@

We have also found a large number of very wide separation ($>1$~pc)
mutually exclusive co-moving pairs, in which the stars are the only co-moving neighbor
of each other and not part of large connected components.
These are most likely remnants of dissolving wide binaries (\citealt{Jiang:2010aa}).
The abundance of highly probable wide separation co-moving pairs
conclusively shows that there is no strict cut-off semi-major axis for wide binary systems
(e.g., \citealt{Wasserman:1987aa}).
The presence of these pairs and similar separation distribution have already
been noticed by \citealt{Shaya:2011aa} using the \project{Hipparcos} data.
If confirmed with radial velocity measurements, this population should still be
relatively young compared to the general disk field population.
Modeling the color-magnitude diagram distribution of these stars can shed some
light on this issue.
If they are remnants of dissolving systems that were born coeval,
the sample of very wide separation co-moving pairs can potentially be used
to measure the recent ($\lesssim 1$~Gyr)
star formation history in the Solar neighborhood.
Co-moving stars with separation less than 1~pc are very promising candidates
for wide binaries. They are found to be pairs of stars of varying stellar types.
Some of these pairs, such as main sequence-red giant or F/G/K-M dwarfs,
will be particularly valuable for testing theoretical
stellar models and calibrating observational measurements of low mass stars.

@@

In what follows, all vectors are column vectors, unless we have transposed them.
A relevant exponential integral solution is
\begin{eqnarray}
  \ln\left[\int\exp(-\frac{1}{2}\,
    \transp{[\vec{x}-\vec{\nu}]} \,
    \inv{\mat{A}} \,
    [\vec{x}-\vec{\nu}] - \Delta) \, \dd \vec{x}\right]
  &=& +\frac{1}{2}\ln ||2\pi\,\mat{A}|| -\Delta
  \quad , \label{eq:gauss-int}
\end{eqnarray}
where $\vec{x}$ and $\vec{\nu}$ are $D$-dimensional vectors, $\mat{A}$ is a
positive definite matrix, $\Delta$ is a scalar, and the integral is over all of
$D$-dimensional $\vec{x}$-space.
To cast our problem in this form, we will need to complete the square of the
exponential argument.
If we equate
\begin{eqnarray}
  \frac{1}{2}\,\transp{[\vec{x}-\vec{\nu}]}\,\inv{\mat{A}}\,[\vec{x}-\vec{\nu}] + \Delta
  &=& \frac{1}{2}\,\transp{\vec{x}}\,\inv{\mat{A}}\,\vec{x} + \transp{\vec{x}}\,\mat{B}\,\vec{b} + C
  \quad ,
\end{eqnarray}
where $\mat{B}\,\vec{b}$ is a $D$-vector, and $C$ is a scalar, then we find
\begin{eqnarray}
  \vec{\nu} &=& -\mat{A}\,\mat{B}\,\vec{b} \label{eq:convnu}
  \\
  \Delta & = & C - \frac{1}{2}\,\transp{\vec{\nu}}\,\inv{\mat{A}}\,\vec{\nu} \label{eq:convDelta}
  \quad .
\end{eqnarray}
We will identify terms in our likelihood functions with $\mat{A}$,
$\mat{B}\,\vec{b}$, and $C$, convert to $\vec{\nu}$ and $\Delta$ and compute
the marginalized likelihood using \eqname~\ref{eq:gauss-int}.

@@

At given distance $r$, the velocity-marginalized likelihood can be computed
analytically using the expressions in Appendix~\ref{sec:appendixA}.
We will start by writing down expressions for the the likelihood multiplied by
the prior pdf for the velocities.
The likelihood for the data (proper motions of the two stars in a pair)
is a Gaussian (\eqname~\ref{eq:likefn}).
In order to simplify our notation, 
we construct a velocity-space data vector $\vec{y}$ as follows:
\begin{equation}
  \vec{y} =
    \transp{\left(
      \begin{array}{c c c c}
        r_i\,\mu_{\alpha,i} &
        r_i\,\mu_{\delta,i} &
        r_j\,\mu_{\alpha,j} &
        r_j\,\mu_{\delta,j}
      \end{array}
    \right)}
\end{equation}
where the subscripts $i,j$ refer to the indices of each star in the pair
and we have multiplied the observables (the proper motions) by the distances
$r_i, r_j$, which is permitted because we are conditioning on the distances.
Fundamentally, our hypothesis 1 model (the stars have the same velocity with a
small difference) is
\begin{equation}
  \vec{y} = \mat{M} \, \vec{v} + \mathrm{noise}
\end{equation}
where now the $4 \times 3$ transformation matrix $\mat{M}$ is a stack of the
transformation matrices for each star computed from the pair of sky positions
and using \eqname~\ref{eq:transformation}.
The noise (in $\vec{y}$) is drawn from a $4 \times 4$ Gaussian with
block-diagonal covariance matrix, $\mat{\Sigma}$, constructed from
the proper motion covariance matrix of each star,
$\mat{C}_i, \mat{C}_j$, and the distances
$r_i, r_j$:
\begin{equation}
  \mat{\Sigma} = \left(
    \begin{array}{c c}
      r_i^2 \, \mat{C}_i & 0 \\
      0 & r_j^2 \, \mat{C}_j
    \end{array}
  \right)
\end{equation}

Given these definitions, the likelihood function for hypothesis 1 is
\begin{eqnarray}
  p(\data \given \vec{v}, r_i, r_j) &=& r_i^2\,r_j^2\,
    \normal(\vec{y} \given \mat{M}\,\vec{v}, \mat{\Sigma}) \\
  \ln p(\data \given \vec{v}, r_i, r_j) &=& 2\,\ln r_i + 2\,\ln r_j
    -\frac{1}{2}\,\ln||2\pi\,\mat{\Sigma}|| \nonumber \\
    && \quad -\frac{1}{2}\,\transp{[\vec{y}-\mat{M}\,\vec{v}]}\,
      \inv{\mat{\Sigma}}\,
      [\vec{y}-\mat{M}\,\vec{v}]
  \quad ,
\end{eqnarray}
where the factor of $r_i^2\,r_j^2$ is the Jacobian from $\vec{y}$
to data ($\transp{(\mu_{\alpha,i}\ \,\mu_{\delta,i}\ \,\mu_{\alpha,j}\ \,\mu_{\delta,j})}$) space.

Now we multiply this likelihood with the velocity prior.
As described in \sectionname~\ref{sec:methods}
we use an isotropic, mixture-of-Gaussians prior on velocity (\eqname~\ref{eq:vprior}).
For simplicity here let us work out the marginalization for one component of
the mixture so that $\vec{v} \sim \normal(\vec{0}, \mat{V}_m)$.
Then,

\begin{eqnarray}
  \ln p(\vec{v}) = -\frac{1}{2}\,\ln||2\pi\,\mat{V}_m||
  -\frac{1}{2} \transp{\vec{v}}\,\inv{\mat{V}_m}\,\vec{v}
\end{eqnarray}
We can identify $\mat{A}$, $\mat{B}\,\vec{b}$, and $C$ in
$\ln p(\vec{v}) + \ln p(\data \given \vec{v}, r_i, r_j)$, and write down
$\vec{\nu}$ and $\Delta$ using \eqname~\ref{eq:convnu} and \ref{eq:convDelta} as

\begin{eqnarray}
  \mat{A} &=& \inv{[\transp{\mat{M}}\,\inv{\mat{\Sigma}}\,\mat{M}+\inv{\mat{V}_m}]}
  \\
  \vec{\nu} &=& -\mat{A}\,\transp{\mat{M}}\,\inv{\mat{\Sigma}}\,\vec{y}
  \\
  \Delta &=& -2\,\ln r_i -2\,\ln r_j
    +\frac{1}{2}\,\ln||2\pi\,\mat{\Sigma}|| +\frac{1}{2}\,\ln||2\pi\,\mat{V}_m|| \nonumber \\
    && \quad +\frac{1}{2}\,\transp{\vec{y}}\,\inv{\mat{\Sigma}}\,\vec{y} -\frac{1}{2}\,\transp{\vec{\nu}}\,\inv{\mat{A}}\,\vec{\nu}
  \quad ,
\end{eqnarray}
which we plug in to \eqname~\ref{eq:gauss-int} to get the marginalized
likelihood conditioned on the two distances $r_i, r_j$.

The marginalized likelihood for the hypothesis 2 model (the stars have
independent velocities) is very similar.
In this case, the marginalized likelihood is a product of two independent
integrals $Q$, composed in the same way as the hypothesis 1 model but now for
each star individually, where
\begin{eqnarray}
  \vec{y} &=&
    \transp{\left(
      \begin{array}{c c}
        r\,\mu_{\alpha} &
        r\,\mu_{\delta}
      \end{array}
    \right)}
  \\
  \mat{\Sigma} &=& r^2 \, \mat{C}
  \\
\end{eqnarray}
and $\mat{M}$ is now the transformation matrix for one star. Then,
\begin{eqnarray}
  \mat{A} &=& \inv{[\transp{\mat{M}}\,\inv{\mat{\Sigma}}\,\mat{M}+\inv{\mat{V}_m}]}
  \\
  \vec{\nu} &=& -\mat{A}\,\transp{\mat{M}}\,\inv{\mat{\Sigma}}\,\vec{y}
  \\
  \Delta &=& -2\,\ln r
    +\frac{1}{2}\,\ln||2\pi\,\mat{\Sigma}|| +\frac{1}{2}\,\ln||2\pi\,\mat{V}_m|| \nonumber \\
    && \quad +\frac{1}{2}\,\transp{\vec{y}}\,\inv{\mat{\Sigma}}\,\vec{y} -\frac{1}{2}\,\transp{\vec{\nu}}\,\inv{\mat{A}}\,\vec{\nu}
  \quad ,
\end{eqnarray}
and
\begin{equation}
  Q = \frac{1}{2}\ln ||2\pi\,\mat{A}|| -\Delta \quad .
\end{equation}
